\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-18 pldi/draft.tex
Update section 3

% For comments
\newcommand{\eat}[1]{}
\newcommand{\TODO}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\todo}[1]{\hl{#1}\xspace}
\newcommand{\fixme}[1]{[{\color{red}#1}]}
\newcommand{\nv}[1]{[{\color{cyan}#1 --- nv}]}
\newcommand{\kk}[1]{[{\color{magenta}#1 --- kk}]}
\newcommand{\km}[1]{[{\color{blue}#1 --- km}]}
\newcommand{\review}[1]{{\color{red}#1}}
\newcommand{\tr}[1]{} %% Text and comments for technical report

@@

\lstset{ %
  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\small\ttfamily,  % the size of the fonts that are used for the code
  upquote=true,
  captionpos=b,                    % sets the caption-position to bottom
% frame=B,                    % adds a frame around the code
  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
  numbersep=2pt,                   % how far the line-numbers are from the code
  numberstyle=\tiny\color{gray},   % the style that is used for the line-numbers
  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
  framerule=0pt,
	xleftmargin=0pt,
	xrightmargin=0pt,
	breakindent=0pt,
  aboveskip=0pt,
  framesep=0pt,
  abovecaptionskip=0pt,
  aboveskip=0pt,
  belowcaptionskip=0pt,
  belowskip=0pt,
  frame=none,
  framexbottommargin=0pt,
  resetmargins=true
}

@@

\begin{table}[t]
\center
\footnotesize
\setlength\tabcolsep{3pt}
\caption{
  \footnotesize{
    \textbf{Parallelizability Classes}.
    Broadly, \unix commands can be broken down into to \todo{three} classes according to their parallelizability properties.
  }
}
\begin{tabular*}{\columnwidth}{l @{\extracolsep{\fill}} llll}
\toprule
Class                    &  Key    & Examples                                    & Coreutils              & POSIX       \\
\midrule
Stateless                & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)           & 22 (21.1\%)          \\  % 22
Parallelizable Pure      & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  \todo{21 (20.1\%)}    & 21 (20.1\%)          \\  % 21
Non-parallelizable Pure  & ~\tnpu  & \tti{sha1sum}                               &  \todo{27 (25.9\%)}    & 27 (25.9\%)          \\  % 27
Side-effectful           & ~\tsid  & \tti{env},  \tti{cp}, \tti{whoami}          &  57 (58.8\%)           & 57 (58.8\%)            \\  % 25
% Irreversible     & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)     & 5  (0.4\%)           \\  % 5
% \midrule
% Shell PL Constructs             &                         &                  \\
% \etc                            &                         &                  \\
\bottomrule
\end{tabular*}
\label{tab:classes}
\end{table}

@@

% \begin{figure}[t]
% \centering
% \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_classes.pdf}}
% \caption{
%   \textbf{\sys's parallelizability classes.}
% }
% \label{fig:example}
% \end{figure}

@@

@@

This is the largest class because the file and file-system abstractions are central to the \unix design and philosophy~\cite{unix}.
In fact, \unix uses the file-system as a proxy to several file-unrelated operations such as access control and device driving. % is grafted upon the file-system.

@@

\subsection{Extensibility}
\label{ext}

\nv{Note command breadth and the impossibility of our task.}
% https://nextbreakpoint.com/posts/article-compile-code-with-docker.html

@@

% FIXME: \nv{TODO: remember to bubble up(me)}
An important characteristic of the \unix shell is its
language-agnostic extensibility~\sx{bg}.  Users composing pipelines
are free to install custom commands available from various sources.
Being general programs, commands are developed in a variety of
languages and are extended over long periods of time.
%% \kk{Text point: This restricts the space of solutions for
%%   distribution of shell scripts}
As a result, any static characterization of shell commands would bs
unsatisfactory and would quickly become obsolete, supporting only a
small subset of the commands needed in practice.  This poses a
challenge to the developers of individual commands---how can they
augment their commands to capture information critical to \sys without
too much effort?
%% \kk{Text point: The characteristics of a good solution}

To address this challenge, \sys arms command developers with a
lightweight, parallelizability annotation language (\todo{PAL}) that
allows devevelopers to express semantic information about a command's
parallelizability.  The PAL can be used by both developers of new
commands, as well as maintainers of existing commands that can express
additions or changes to the command's implmentation or
interface. \todo{The two main challenges that PAL has to address is
  supporting commands with mutiple inputs and outputs, and handle
  command-line options.}

\heading{Command Inputs and Outputs}

Together with a command's parallelizability class (i.e. \sta, \pur,
etc), \sys needs to know certain information about a command's inputs
and outputs in order to construct a dataflow graph. More precisely,
given a command and its arguments, \sys needs to know its inputs and
outputs to connect commands in the dataflow graph with edges, as well
as the order in which a command consumes its input; this is important
for \sys to know how to order parallel instances of a command after it
has performed any \todo{optimizations}. Consider the example of
\ttt{grep "foo" f1 f2}, which can be parallelized as follows if we
know that \ttt{grep} first reads from \ttt{f1} and then from \ttt{f2}.

@@

\heading{Command-line Options}
As described earlier~\sx{bg}, com\-mand-line options and environment
variables can affect the operation of a command.  Options are a
particularly popular way to control a command, directly affecting a
command's parallelizability classification.  That is, commands are
assigned to default classes, and options can change the class in which
they belong.  As a simple example, consider \ttt{cat}.  By default
\ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it
has to keep track of a counter and output the line number with each line.

@@

%% This is actually wrong. sort -h is not stateless since if called several times it produces the same output many times.

@@

\tr{
\begin{figure}
  \centering
  \begin{grammar}
    <option> ::= `-' <string>

    <category> ::= `stateless' | `pure' | ...

    <maybe-int> ::= | <int>

    <arg> ::= `args[' <int> `]'

    <args> ::= <arg>
    \alt `args[' <maybe-int> `:' <maybe-int> `]'

    <input> ::= `stdin' | <args>

    <inputs> ::= <input>
    \alt <input> `,' <inputs>

    <output> ::= `stdout' | <arg>

    <option-pred> ::= <option>
    \alt `value' <option> = <string>
    \alt `not' <option-pred>
    \alt <option-pred> `or' <option-pred>
    \alt <option-pred> `and' <option-pred>

    <assignment> ::= `(' <category>, `[' <inputs> `]' , <output> `)'

    <predicate> ::= <option-pred> `=>' <assignment>

    <pred-list> ::= `|' <predicate> <pred-list>
    \alt `|' `otherwise' `=>' <assignment>

    <command> ::= <name> `\{' <pred-list> `\}'

    <command-list> ::= <command>
    \alt <command> <command-list>
  \end{grammar}
  \caption{
\textbf{Parallelizability description language.}
The DSL captures important information regarding the parallelizability of a command.
\vspace{-10pt}
}
  \label{fig:dsl}
\end{figure}
}

@@

\heading{Parallelizability Annotation Language}

To illustrate how the proposed PAL addresses the above challenges, we
will follow the example of the annotation record for \ttt{comm}, a
command that performs a ``join'' on its two inputs. When \ttt{comm} is
invoked without any option, it produces a three-column output. The
first column contains lines that are unique to the first file, the
second column contains lines that are unique to the second file, and
the last column contains lines that exist in both files. A user can
invoke \ttt{comm} with options \ttt{-1}, \ttt{-2}, or \ttt{-3} to
suppress some of its output columns.
%
\begin{lstlisting}[float=h, numbers=none,caption={PAL record for \ttt{comm}},label={lst:pal-record-comm}]
  comm {
    | -1 and -3 =>
      (stateless, [args[1]], [stdout])
    | -2 and -3 =>
      (stateless, [args[0]], [stdout])
    | otherwise =>
      (pure, [args[0], args[1]], [stdout])
  }
\end{lstlisting}

The PAL annotation record for \ttt{comm} is shown in
\cref{lst:pal-record-comm}. A record contains a set of clauses, each
one identified by a predicate on the commands options. Each clause
maps to an assignment for the command, indicating its class and the
sequence of its inputs and outputs. \todo{PAL assumes} that arguments
of a command can be separated to options and file arguments (checking
if they start with a \ttt{-}~\footnote{Note that the single character
  \ttt{-} is not an option as it is most often used to designate that
  a command reads from its standard input}).

In the record for \ttt{comm} the first two clauses describe \ttt{comm}
invocations with options \ttt{-1} (or \ttt{-2}) and \ttt{-3}. In the
first case, since the first and third columns are suppressed,
\ttt{comm} can be considered \sta by regarding its first file argument
a static ``configuration'' input. \kk{Does this need more
  explanation?} The general case (that includes \ttt{comm} invocations
without any options) is captured by the third clause. There,
\ttt{comm} is classified as \pur, since it needs to wait until it
reads both files completely before it can output its results. Its
inputs are its first and second arguments (excluding the options), and
its output is produced in \ttt{stdout}.

@@

\kk{We could add the grammar of PAL in an appendix if we want}

\tr{The complete syntax of the PAL is shown in (Fig.~\ref{fig:dsl})}
%% \kk{I
%%   think that the language definition should be in a technical report,
%%   and here we should only give intuition and the \ttt{comm} example}
%% is designed to classify commands into multiple categories given their
%% options.

@@

\heading{Parallelizable Pure Commands}

In addition to the commands in \sta that can be parallelized in a
straightforward way, there also exists a subclass of commands in \pur
that can be parallelized using divide-and-conquer techniques (as
mentioned in \cref{cmd}). \kk{Maybe show an example of map-fold for
  sort or wc. Either here or in \cref{cmd}.} In order to support the
parallelization of arbitrary commands in this category, \sys allows
command developers to provide custom map and fold steps for them,
where the map is in \sta and the fold in \pur, and their composition
produces the same output with the initial command. \sys can then
replace occurrences of these commands with their map and fold parts,
enabling their parallelization. \todo{We have defined these primitives
  for most of the parallelizable pure commands in coreutils, providing
  a foundation for this effort.}

@@

>>>>

% For comments
\newcommand{\eat}[1]{}
\newcommand{\TODO}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\todo}[1]{\hl{#1}\xspace}
\newcommand{\fixme}[1]{[{\color{red}#1}]}
\newcommand{\nv}[1]{[{\color{cyan}nv: #1}]}
\newcommand{\kk}[1]{[{\color{magenta}kk: #1}]}
\newcommand{\km}[1]{[{\color{blue}km: #1}]}
\newcommand{\review}[1]{{\color{red}#1}}
\newcommand{\tr}[1]{} %% Text and comments for technical report

@@

\lstset{ %
  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\normalsize\ttfamily,  % the size of the fonts that are used for the code
  upquote=true,
  captionpos=b,                    % sets the caption-position to bottom
% frame=B,                    % adds a frame around the code
  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
  numbersep=2pt,                   % how far the line-numbers are from the code
  numberstyle=\tiny\color{gray},   % the style that is used for the line-numbers
  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
  framerule=0pt,
	xleftmargin=0pt,
	xrightmargin=0pt,
	breakindent=0pt,
  aboveskip=0pt,
  framesep=0pt,
  abovecaptionskip=0pt,
  aboveskip=0pt,
  belowcaptionskip=0pt,
  belowskip=0pt,
  frame=none,
  framexbottommargin=0pt,
  resetmargins=true
}

@@

\begin{table}[t]
\center
\footnotesize
\setlength\tabcolsep{3pt}
\caption{
  \footnotesize{
    \textbf{Parallelizability Classes}.
    Broadly, \unix commands can be broken down into to \todo{four} classes according to their parallelizability properties.
  }
}
\begin{tabular*}{\columnwidth}{l @{\extracolsep{\fill}} llll}
\toprule
Class                    &  Key    & Examples                                    & Coreutils              & POSIX       \\
\midrule
Stateless                & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)           & 22 (21.1\%)          \\  % 22
Parallelizable Pure      & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  \todo{21 (20.1\%)}    & 21 (20.1\%)          \\  % 21
Non-parallelizable Pure  & ~\tnpu  & \tti{sha1sum}                               &  \todo{27 (25.9\%)}    & 27 (25.9\%)          \\  % 27
Side-effectful           & ~\tsid  & \tti{env},  \tti{cp}, \tti{whoami}          &  57 (58.8\%)           & 57 (58.8\%)            \\  % 25
% Irreversible     & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{reboot}      &  5  (0.4\%)     & 5  (0.4\%)           \\  % 5
% \midrule
% Shell PL Constructs             &                         &                  \\
% \etc                            &                         &                  \\
\bottomrule
\end{tabular*}
\label{tab:classes}
\end{table}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.20\textwidth]{\detokenize{./figs/dish_classes.pdf}}
\caption{
  \textbf{Hierarchy of parallelizability classes.}
  Parallelizability classes form a hierarchy from the easiest to parallelize (left) to the most difficult (right).
  \nv{The figure is incorrect---it's actually a lattice.}
}
\vspace{-15pt}
\label{fig:example}
\end{figure}

@@

\kk{Maybe show an example of map-fold for sort or wc}

@@

This is the largest class because it includes commands related to the file-system---a central abstractions to the \unix design and philosophy~\cite{unix}.
In fact, \unix uses the file-system as a proxy to several file-unrelated operations such as access control and device driving. % is grafted upon the file-system.

@@

\subsection{Extensibility Annotations}
\label{ext}

\nv{Note command breadth/impossibility of our task.}
% https://nextbreakpoint.com/posts/article-compile-code-with-docker.html

@@

% % FIXME: \nv{TODO: remember to bubble up(me)}
% An important characteristic of the \unix shell is its
% language-agnostic extensibility~\sx{bg}.  Users composing pipelines
% are free to install custom commands available from various sources.
% Being general programs, commands are developed in a variety of
% languages and are extended over long periods of time.
% %% \kk{Text point: This restricts the space of solutions for
% %%   distribution of shell scripts}
% As a result, any static characterization of shell commands would bs
% unsatisfactory and would quickly become obsolete, supporting only a
% small subset of the commands needed in practice.  This poses a
% challenge to the developers of individual commands---how can they
% augment their commands to capture information critical to \sys without
% too much effort?
% %% \kk{Text point: The characteristics of a good solution}

To address the challenge of language-agnostic extensibility~\sx{bg}, \sys allows communicating several key details about command parallelizability through lightweight annotations.
These annotations can be used by both developers of new commands---which includes users developing their own scripts---as well as developers maintaining existing commands.
The latter can express additions or changes to the command's implmentation or interface, important as commands are maintained or extended over long periods of time.

\heading{Key Concerns}
\sys's annotations focused on three crucial concerns:
  (i) a command's parallelizability class,
  (ii) for commands that support multiple inputs, the characteristics of input consumption, and
  (iii) for commands with multiple flags (options), the handling of \todo{competing sets} of flags.
As the first concern was discussed extensively in the previous section, we now focus on the latter two.\footnote{
  \nv{Note to self: until now we have been talking only about parallelizability, and now we introduce two concerns that haven't been mentioned before.}
}

To be able to accurately construct the DFG representation, \sys needs to know certain information about a command's inputs and outputs.
There are couple of reasons for this.
One reason is due to \sys's DFG representation, which connects commands with each other;
  to perform this connection correctly \sys needs to know the outputs of a command to correctly connect them with the inputs of the next.
A second reason is due to ordering:
  as some commands consume their inputs in a certain order, this order needs to be maintained in the parallel version of the program.
For example, consider the parallel version of the expression \ttt{grep "foo" f1 f2}:

@@

\noindent
This is correct only because \sys knows that \ttt{grep} in the sequential program reads from \ttt{f1} first and from \ttt{f2} second.

Command flags (options) are a particularly popular way to control a command's execution, directly affecting its parallelizability classification.
Commands are thus assigned a default parallelizability class, which is then refined by the set of flags the command uses.
For example, \ttt{cat} defaults to \sta, but with \ttt{-n} it jumps into \pur because it has to keep track of a counter and print it along with each line.

As parallelizability classes form a hierarchy from most parallelizable to least parallelizable~\sx{cmd}, a command is classified by the class of its least parallelizable flag.
For example, if a custom \ttt{trace-sort} command is invoked with flags \ttt{r}, \ttt{n}, and \ttt{k2} that are in \pur and a flag \ttt{d} that writes debugging output to a file, it ends up in \sta.

@@

%% This is actually wrong. sort -h is not stateless since if called several times it produces the same output many times.
%% nv: the above does not mean it is not stateless; if apply `map` multiple times on an input I will get combined results.

@@

@@

\heading{Example Annotations}
A record contains a set of clauses, each one identified by a predicate on the commands options.
Each clause maps to an assignment for the command, indicating its class and the sequence of its inputs and outputs.
Annotations treat flag arguments differently from file arguments, by checking if they start with a \ttt{-}.
% \footnote{
%   Note that the single character \ttt{-} is not an option as it is most often used to designate that a command reads from its standard input.
%   \nv{
%     There are options of of the form --f "x; y", and BSD options do not take \ttt{-} (like \ttt{ps aux}).
%     I am also not sure how important this last sentence is, as it did not answer a question I had---I expected arguments to be anything: if soneome wants to put ``pizza'' the DSL should not disallow it.
%   }
% }
The grammar of the language for specifying annotations is shown in Appendix~\ref{alg}.

For example, consider \ttt{comm}---a command that performs a join-like operation on its two inputs to identify common elements.
When \ttt{comm} is invoked without any flags, it produces a three-column output:
  the first column contains lines that are unique to the first input, the second column contains lines that are unique to the second input, and the last column contains lines that exist in both inputs.
A user can invoke \ttt{comm} with any combination of flags \ttt{-1}, \ttt{-2}, or \ttt{-3} to suppress the corresponding output columns.
Its annotation record is shown below:
%
\begin{lstlisting}[language=sh, numbers=none]
comm {
  | -1 /\ -3 =>
    (stateless, [args[1]], [stdout])
  | -2 /\ -3 =>
    (stateless, [args[0]], [stdout])
  | _  =>
    (pure, [args[0], args[1]], [stdout])
}
\end{lstlisting}

\noindent
The first two clauses describe \ttt{comm} invocations with options \ttt{-1} (or \ttt{-2}) and \ttt{-3}.
In the first case, since the first and third columns are suppressed, \ttt{comm} can be considered in \sta by regarding its first file argument as a static ``configuration'' input;
  this input accompanies every replicated instance of the command executing in parallel.
The general case, which includes \ttt{comm} invocations without any options, is captured by the third clause.
There, \ttt{comm} is classified as \pur, since it needs to wait until it reads both files completely before it can output its results.
Its inputs are its first and second file arguments (\ie excluding any options), and its output is produced in \ttt{stdout}.

@@

%% \kk{I
%%   think that the language definition should be in a technical report,
%%   and here we should only give intuition and the \ttt{comm} example}
%% is designed to classify commands into multiple categories given their
%% options.

@@

\heading{Providing Combinators}
For most commands, the annotations are enough to allow parallelization: 
  commands either fall under \sta, having straightforward combiners, or fall under a subset of \pur that can be parallelized using divide-and-conquer techniques~\sx{cmd}.
Unfortunately, there are commands with less trivial tactics---after all, \unix commands are Turing-complete programs and failing to plan for this generality would inhibit \sys's usefulness.

To support the parallelization of arbitrary commands in \pur, \sys allows command developers to provide custom $map$ and $fold$ functions.
In line with \unix philosophy, these functions can be written in any language as long as they conform to a few invariants:
  (i) $map$ in \sta and $fold$ is in \pur,
  (ii) $map$ can consume (or extend) the output of the original command and  $fold$ can consume (and combine) the results of multiple $map$ invocations, and
  (iii) their composition produces the same output with the original command.
If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their $map$ and $fold$ equivalents, enabling their parallelization.

\sys defines the combiners for most of the POSIX and Coreutils commands in \pur, which operate as both \sys's standard library and exemplar for community attempts to tackle other commands.

@@

\appendix

\newpage

\section{Annotation Language Grammar}
\label{alg}

% \begin{figure}
%   \centering
%   \begin{grammar}
%     <option> ::= `-' <string>
% 
%     <category> ::= `stateless' | `pure' | ...
% 
%     <maybe-int> ::= | <int>
% 
%     <arg> ::= `args[' <int> `]'
% 
%     <args> ::= <arg>
%     \alt `args[' <maybe-int> `:' <maybe-int> `]'
% 
%     <input> ::= `stdin' | <args>
% 
%     <inputs> ::= <input>
%     \alt <input> `,' <inputs>
% 
%     <output> ::= `stdout' | <arg>
% 
%     <option-pred> ::= <option>
%     \alt `value' <option> = <string>
%     \alt `not' <option-pred>
%     \alt <option-pred> `or' <option-pred>
%     \alt <option-pred> `and' <option-pred>
% 
%     <assignment> ::= `(' <category>, `[' <inputs> `]' , <output> `)'
% 
%     <predicate> ::= <option-pred> `=>' <assignment>
% 
%     <pred-list> ::= `|' <predicate> <pred-list>
%     \alt `|' `otherwise' `=>' <assignment>
% 
%     <command> ::= <name> `\{' <pred-list> `\}'
% 
%     <command-list> ::= <command>
%     \alt <command> <command-list>
%   \end{grammar}
%   \caption{
%   \textbf{Parallelizability description language.}
%     The DSL captures important information regarding the parallelizability of a command.
%   }
%   \label{fig:dsl}
% \end{figure}
